---
title: "Global_fishing_watch_workflow_envpreds"
author: "Théophile L. Mouton"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-location: right
    css: custom.css
    output-file: "Global_Fishing_Watch_workflow_envpreds.html"
    self-contained: true
    code-fold: true
    code-tools: true
editor: visual
execute:
  warning: false
params:
  printlabels: TRUE
---

# Combining AIS and Sentinel-1 SAR fishing effort data from Global Fishing Watch

## Open R libraries

```{r}
# Load required libraries
library(tidyverse)
library(maps)
library(ggplot2)
library(data.table)
library(gridExtra)
library(raster)
library(sf)
library(viridis)
library(scales)
library(dplyr)
library(randomForest)
library(caret)
library(pdp)
library(knitr)
library(kableExtra)
library(future)
library(spdep)
library(ncf)
library(blockCV)
library(doParallel)
library(sp)
```

## AIS data

Data from Kroodsma et al. (2018) Science, accessible at: [Global Fishing Watch Data Download Portal](https://globalfishingwatch.org/data-download/datasets/public-fishing-effort).

The data is accessible up to the end of the year 2020, we used four entire years of data (2017, 2018, 2019, 2020) to match SAR data records.

```{r}
# Set the path to the 2017-2020 folder

#path <- "Data/AIS Fishing Effort 2017-2020"

# List all CSV files in the folder
#AIS_csv_files <- list.files(path = here::here(path), pattern = "*.csv", full.names = TRUE, recursive = TRUE)

# Read all CSV files and combine them into a single data frame
#AIS_fishing <- AIS_csv_files %>%
#  map_df(~read_csv(.))

load(here::here("Data","AIS_fishing.Rdata"))

# Aggregate fishing hours by latitude and longitude
aggregated_AIS_fishing <- AIS_fishing %>%
  group_by(cell_ll_lat, cell_ll_lon) %>%
  summarise(total_fishing_hours = sum(fishing_hours, na.rm = TRUE)) %>%
  ungroup() %>%
  # Remove any cells with zero or negative fishing hours
  filter(total_fishing_hours > 0)

# Function to standardize coordinates to 0.1 degree resolution
standardize_coords <- function(lon, lat) {
  list(
    lon_std = floor(lon * 10) / 10,
    lat_std = floor(lat * 10) / 10
  )
}

# Standardize and aggregate AIS data
AIS_data_std <- aggregated_AIS_fishing %>%
  mutate(coords = map2(cell_ll_lon, cell_ll_lat, standardize_coords)) %>%
  mutate(
    lon_std = map_dbl(coords, ~ .x$lon_std),
    lat_std = map_dbl(coords, ~ .x$lat_std)
  ) %>%
  group_by(lon_std, lat_std) %>%
  summarise(total_fishing_hours = sum(total_fishing_hours, na.rm = TRUE), .groups = "drop")

# Create the world map
world_map <- map_data("world")

# Create the plot
ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(long = long, lat = lat, map_id = region),
           color = "black", fill = "lightgray", size = 0.1) +
  geom_tile(data = AIS_data_std, 
            aes(x = lon_std, y = lat_std, fill = total_fishing_hours)) +
  scale_fill_viridis(
    option = "inferno",
    direction = -1,
    trans = "log1p",
    name = "AIS Fishing Effort (hours; 2017-2020)", 
    breaks = c(0, 1, 10, 100, 1000, 10000, 100000),
    labels = scales::comma,
    guide = guide_colorbar(barwidth = 20, barheight = 0.5, title.position = "top", title.hjust = 0.5)
  ) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = NULL,
       x = "Longitude", y = "Latitude") +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.margin = ggplot2::margin(t = 20, r = 0, b = 0, l = 0),
    legend.title = element_text(margin = ggplot2::margin(b = 10))
  )
```

## Sentinel-1 SAR data

Data from Paolo et al. 2024 Nature, accessible at: [Global Fishing Watch SAR Vessel Detections](https://globalfishingwatch.org/data-download/datasets/public-sar-vessel-detections:v20231026)

The data is accessible from 2017 to today, we used four entire years of data (2017, 2018, 2019, 2020) to match AIS records.

```{r}
# Set the path to the 2016 folder
#path <- "Data/SAR Vessel detections 2017-2020"

# List all CSV files in the folder
#SAR_csv_files <- list.files(path = here::here(path), pattern = "*.csv", full.names = TRUE)

# Read all CSV files and combine them into a single data frame
#SAR_fishing <- SAR_csv_files %>%
#  map_df(~read_csv(.))

load(here::here("Data","SAR_fishing.Rdata"))

# Aggregate fishing hours by latitude and longitude
aggregated_SAR_fishing <- SAR_fishing %>%
  mutate(
    lat_rounded = round(lat, digits = 2),
    lon_rounded = round(lon, digits = 2)
  ) %>%
  group_by(lat_rounded, lon_rounded) %>%
  filter(fishing_score >= 0.9) %>%
  summarise(
    total_presence_score = sum(presence_score, na.rm = TRUE),
    avg_fishing_score = mean(fishing_score, na.rm = TRUE),
    count = n()
  ) %>%
  mutate(total_presence_score = round(total_presence_score, digits = 0)) %>%
  ungroup()

# Standardize and aggregate SAR data
SAR_data_std <- aggregated_SAR_fishing %>%
  mutate(coords = map2(lon_rounded, lat_rounded, standardize_coords)) %>%
  mutate(
    lon_std = map_dbl(coords, ~ .x$lon_std),
    lat_std = map_dbl(coords, ~ .x$lat_std)
  ) %>%
  group_by(lon_std, lat_std) %>%
  summarise(total_presence_score = sum(total_presence_score, na.rm = TRUE), .groups = "drop")

# Create the world map
world_map <- map_data("world")

# Create the plot
ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(long = long, lat = lat, map_id = region),
           color = "black", fill = "lightgray", size = 0.1) +
  geom_tile(data = SAR_data_std, 
            aes(x = lon_std, y = lat_std, fill = total_presence_score)) +
  scale_fill_viridis(
    option = "inferno",
    direction = -1,
    trans = "log1p",
    name = "Fishing vessel detections (2017-2020)", 
    breaks = c(0, 1, 10, 100, 1000, 10000),
    labels = scales::comma,
    guide = guide_colorbar(barwidth = 20, barheight = 0.5, title.position = "top", title.hjust = 0.5)
  ) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = NULL,
       x = "Longitude", y = "Latitude") +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.margin = ggplot2::margin(t = 20, r = 0, b = 0, l = 0),
    legend.title = element_text(margin = ggplot2::margin(b = 10))
  )
```

## Compare AIS, and SAR detection locations

Identifying grid cells with only AIS, only SAR detections or both data types.

```{r}
# Merge the datasets
combined_data <- full_join(
  AIS_data_std,
  SAR_data_std,
  by = c("lon_std", "lat_std")
)

# Categorize each cell
combined_data <- combined_data %>%
  mutate(category = case_when(
    total_fishing_hours > 0 & total_presence_score > 0 ~ "Both AIS and SAR",
    total_fishing_hours > 0 & (is.na(total_presence_score) | total_presence_score == 0) ~ "Only AIS",
    (is.na(total_fishing_hours) | total_fishing_hours == 0) & total_presence_score > 0 ~ "Only SAR",
    TRUE ~ "No fishing detected"
  ))

# Create the world map
world_map <- map_data("world")

# Create the plot
world_plot <- ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(long = long, lat = lat, map_id = region),
           color = "black", fill = "lightgray", size = 0.1) +
  geom_tile(data = combined_data, 
            aes(x = lon_std, y = lat_std, fill = category)) +
  scale_fill_manual(
    values = c("Both AIS and SAR" = "purple", 
               "Only AIS" = "blue", 
               "Only SAR" = "red", 
               "No fishing detected" = "white"),
    name = "Fishing Data Source",
    guide = guide_legend(title.position = "top", title.hjust = 0.5)
  ) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = "Global Fishing Detection",
       subtitle = "Comparison of AIS (2017-2020) and SAR (2017-2020) data at 0.1-degree resolution",
       x = "Longitude", y = "Latitude") +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.margin = ggplot2::margin(t = 20, r = 0, b = 0, l = 0),
    legend.title = element_text(margin = ggplot2::margin(b = 10))
  )

# Display the plot
print(world_plot)

# Get summary statistics
summary_stats <- combined_data %>% 
  count(category) %>% 
  mutate(percentage = n / sum(n) * 100) %>%
  rename(`Number of cells` = n) %>%
  mutate(percentage = round(percentage, 2))

# Create and display the table
kable(summary_stats, 
      format = "html", 
      col.names = c("Category", "Number of cells", "Percentage (%)"),
      caption = "Summary Statistics of Data Categories") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, 
                position = "center") %>%
  column_spec(2, width = "150px") %>%
  column_spec(3, width = "150px")
```

## Random forest model

Predicting fishing hours in areas with only SAR data at a 0.1 degree resolution. Using a random forest model of \>160,000 associations between SAR vessel detections and AIS fishing hours globally, geographical coordinates, distance to ports, distance to shore and bathymetry.

```{r}
# Open the saved adjusted rasters
Ports_adjusted <- raster(here::here("Data", "Environmental data layers", "distance-from-port-0.1deg-adjusted.tif"))
Shore_adjusted <- raster(here::here("Data", "Environmental data layers", "distance-from-shore-0.1deg-adjusted.tif"))
Bathy_adjusted <- raster(here::here("Data", "Environmental data layers", "bathymetry-0.1deg-adjusted.tif"))

# Stack the resampled rasters
raster_stack <- stack(Shore_adjusted, Ports_adjusted, Bathy_adjusted)

# Convert the stack to a dataframe
raster_df <- as.data.frame(raster_stack, xy = TRUE)

# Rename the columns
names(raster_df) <- c("x", "y", "dist_shore", "dist_ports", "bathy")

# Remove NA values if desired
raster_df <- na.omit(raster_df)

# Convert to data.table for efficiency
setDT(raster_df)

# Round x and y to 1 decimal place for consistency
raster_df[, `:=`(
  lon_std = round(x, digits = 1),
  lat_std = round(y, digits = 1)
)]

# Now proceed with the join and model training
load(here::here("data","combined_data_O1deg.Rdata"))
setDT(combined_data_01deg)

# Perform the join using data.table
combined_data_with_rasters <- raster_df[combined_data_01deg, 
                                        on = .(lon_std, lat_std), 
                                        nomatch = 0]

# Convert back to dataframe if needed
combined_data_with_rasters <- as.data.frame(combined_data_with_rasters)

# Prepare the training data
training_data <- combined_data_with_rasters %>%
  filter(has_AIS & has_SAR) %>%
  dplyr::select(total_fishing_hours, total_presence_score, lon_std, lat_std, dist_shore, dist_ports, bathy) %>%
  na.omit()

# Train the random forest model with timing
set.seed(123)  # for reproducibility
#rf_timing <- system.time({
#  rf_model_env <- randomForest(
#    total_fishing_hours ~ total_presence_score + lon_std + lat_std + dist_shore + dist_ports + bathy,
#    data = training_data,
#    ntree = 500,
#    importance = TRUE
#  )
#})

#cat("Random Forest model training time:", rf_timing["elapsed"], "seconds\n")

# Save the new model
#save(rf_model_env, file = "rf_model_01deg_with_rasters.Rdata")
load(here::here("rf_model_01deg_with_rasters.Rdata"))

# Prepare the prediction data
prediction_data <- combined_data_with_rasters %>%
  dplyr::select(total_presence_score, lon_std, lat_std, dist_shore, dist_ports, bathy)

# Make predictions
predictions <- predict(rf_model_env, newdata = prediction_data)

# Add predictions to the original dataset
combined_data_01deg <- combined_data_01deg %>%
  mutate(
    predicted_fishing_hours = case_when(
      has_AIS ~ total_fishing_hours,
      has_SAR ~ predictions[match(paste(lon_std, lat_std), paste(prediction_data$lon_std, prediction_data$lat_std))],
      TRUE ~ 0
    )
  )

# Create the world map
world_map <- map_data("world")

# Map of predicted fishing hours only 
# Prepare the data for the map
map_data <- combined_data_01deg %>%
  filter(!has_AIS & has_SAR) %>%
  dplyr::select(lon_std, lat_std, predicted_fishing_hours)

# Function to create map for a specific region
create_region_map <- function(data, world_map, lon_range, lat_range, title) {
  ggplot() +
    geom_map(data = world_map, map = world_map,
             aes(long, lat, map_id = region),
             color = "darkgray", fill = "lightgray", size = 0.1) +
    geom_tile(data = data, 
              aes(x = lon_std, y = lat_std, fill = predicted_fishing_hours)) +
    scale_fill_viridis(
      option = "inferno",
      direction = -1,
      trans = "log1p",
      name = "Predicted fishing hours (2017-2020)", 
      breaks = c(0, 1, 10, 100, 1000, 10000, 100000, 1000000),
      labels = scales::comma,
      guide = guide_colorbar(barwidth = 20, barheight = 0.5, title.position = "top", title.hjust = 0.5)
    ) +
    coord_fixed(1.3, xlim = lon_range, ylim = lat_range) +
    theme_minimal() +
    labs(title = title,
         subtitle = "0.1 degree resolution",
         x = "Longitude", y = "Latitude") +
    theme(
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "vertical",
      legend.margin = ggplot2::margin(t = 20, r = 0, b = 0, l = 0),
      legend.title = element_text(margin = ggplot2::margin(b = 10))
    )
}

# Global map
predicted_SAR_only_plot <- create_region_map(map_data, world_map, c(-180, 180), c(-90, 90), "Predicted Fishing Hours in Areas with Only SAR Detections")

# Caribbean map
caribbean_map <- create_region_map(map_data, world_map, c(-100, -50), c(0, 40), "Predicted Fishing Hours in the Caribbean")

# Northwestern Indian Ocean to Western European waters map
indian_european_map <- create_region_map(map_data, world_map, c(-20, 80), c(0, 70), "Predicted Fishing Hours from Northern Indian Ocean \nto Eastern Atlantic")

# Asia map
asia_map <- create_region_map(map_data, world_map, c(60, 180), c(-20, 60), "Predicted Fishing Hours in Asia")

# Print the maps
print(predicted_SAR_only_plot)
print(caribbean_map)
print(indian_european_map)
print(asia_map)

#Map of both original and predicted AIS fishing hours 
# Visualize the results
predicted_plot <- ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(long, lat, map_id = region),
           color = "black", fill = "lightgray", size = 0.1) +
  geom_tile(data = combined_data_01deg, 
            aes(x = lon_std, y = lat_std, fill = predicted_fishing_hours)) +
  scale_fill_viridis(
    option = "inferno",
    direction = -1,
    trans = "log1p",
    name = "AIS Fishing Effort (hours; 2017-2020)", 
    breaks = c(0, 1, 10, 100, 1000, 10000, 100000, 1000000),
    labels = scales::comma,
    guide = guide_colorbar(barwidth = 20, barheight = 0.5, title.position = "top", title.hjust = 0.5)
  ) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = "Global Fishing Hours (0.1 degree resolution)",
       subtitle = "Based on AIS data and Random Forest predictions from SAR data",
       x = "Longitude", y = "Latitude") +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.margin = ggplot2::margin(t = 20, r = 0, b = 0, l = 0),
    legend.title = element_text(margin = ggplot2::margin(b = 10))
  )

print(predicted_plot)
#Aggregate data to 0.5 degree resolution 
# First, round the coordinates to the nearest 0.5 degree
combined_data_05deg <- combined_data_01deg %>%
  mutate(
    lon_05deg = round(lon_std * 2) / 2,
    lat_05deg = round(lat_std * 2) / 2
  )

# Aggregate the data
aggregated_data <- combined_data_05deg %>%
  group_by(lon_05deg, lat_05deg) %>%
  summarise(
    predicted_fishing_hours = sum(predicted_fishing_hours, na.rm = TRUE),
    .groups = 'drop'
  )

save(aggregated_data, file=here::here("Predicted_Fishing_Hours_05Deg.Rdata"))

# Create the map
predicted_plot_05deg <- ggplot() +
  geom_map(data = world_map, map = world_map,
           aes(long, lat, map_id = region),
           color = "black", fill = "lightgray", size = 0.1) +
  geom_tile(data = aggregated_data, 
            aes(x = lon_05deg, y = lat_05deg, fill = predicted_fishing_hours)) +
  scale_fill_viridis(
    option = "inferno",
    direction = -1,
    trans = "log1p",
    name = "AIS Fishing Effort (hours; 2017-2020)", 
    breaks = c(0, 1, 10, 100, 1000, 10000, 100000, 1000000),
    labels = scales::comma,
    guide = guide_colorbar(barwidth = 20, barheight = 0.5, title.position = "top", title.hjust = 0.5)
  ) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = "Global Fishing Hours (0.5 degree resolution)",
       subtitle = "Based on AIS data and Random Forest predictions from SAR data",
       x = "Longitude", y = "Latitude") +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.margin = ggplot2::margin(t = 20, r = 0, b = 0, l = 0),
    legend.title = element_text(margin = ggplot2::margin(b = 10))
  )

# Print the map
print(predicted_plot_05deg)

# Evaluate the model
evaluate_model <- function(model, data, log_target = FALSE) {
  predictions <- predict(model, newdata = data)
  if (log_target) {
    predictions <- 10^predictions - 1
  }
  
  actual <- data$total_fishing_hours
  
  # Basic Error Metrics
  mae <- mean(abs(actual - predictions), na.rm = TRUE)
  rmse <- sqrt(mean((actual - predictions)^2, na.rm = TRUE))
  mape <- mean(abs((actual - predictions) / actual) * 100, na.rm = TRUE)
  medae <- median(abs(actual - predictions), na.rm = TRUE)
  
  # R-squared (matching randomForest's % Var explained)
  r_squared <- model$rsq[length(model$rsq)]
  
  # Adjusted R-squared
  n <- length(actual)
  p <- length(model$forest$independent.variable.names) # Number of predictors
  adj_r_squared <- 1 - ((1 - r_squared) * (n - 1) / (n - p - 1))
  
  # Residual Analysis
  residuals <- actual - predictions
  mean_residual <- mean(residuals, na.rm = TRUE)
  sd_residual <- sd(residuals, na.rm = TRUE)
  
  # Feature Importance (for Random Forest)
  feature_importance <- importance(model)
  
  return(list(
    "Mean Absolute Error" = mae,
    "Root Mean Squared Error" = rmse,
    "Mean Absolute Percentage Error" = mape,
    "Median Absolute Error" = medae,
    "R-Squared" = r_squared,
    "Adjusted R-Squared" = adj_r_squared,
    "Mean of Residuals" = mean_residual,
    "Standard Deviation of Residuals" = sd_residual,
    "Feature Importance" = feature_importance
  ))
}

validation_data <- combined_data_with_rasters %>%
  mutate(
    data_category = case_when(
      has_AIS & has_SAR ~ "Both AIS and SAR",
      has_AIS & !has_SAR ~ "Only AIS",
      !has_AIS & has_SAR ~ "Only SAR",
      TRUE ~ "No fishing detected"
    )
  )

# Evaluate all models
validation_data <- validation_data %>% filter(data_category == "Both AIS and SAR")
results_rf_env <- evaluate_model(rf_model_env, validation_data)

# Create a dataframe for the table
results_table <- data.frame(
  Metric = c("Mean Absolute Error", "Root Mean Squared Error", "Mean Absolute Percentage Error",
             "Median Absolute Error", "R-Squared", "Adjusted R-Squared",
             "Mean of Residuals", "Standard Deviation of Residuals"),
  Value = round(c(results_rf_env$`Mean Absolute Error`,
                  results_rf_env$`Root Mean Squared Error`,
                  results_rf_env$`Mean Absolute Percentage Error`,
                  results_rf_env$`Median Absolute Error`,
                  results_rf_env$`R-Squared`,
                  results_rf_env$`Adjusted R-Squared`,
                  results_rf_env$`Mean of Residuals`,
                  results_rf_env$`Standard Deviation of Residuals`),
                2)  # Round to 2 decimal places
)

# Create and display the table
kable(results_table, format = "html", digits = 4, caption = "Model Evaluation Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, position = "center")

# For feature importance, create a separate table
feature_importance <- as.data.frame(results_rf_env$`Feature Importance`)
feature_importance$Feature <- rownames(feature_importance)
feature_importance <- feature_importance[, c("Feature", "%IncMSE", "IncNodePurity")]
colnames(feature_importance) <- c("Feature", "%IncMSE", "IncNodePurity")

# Sort the feature importance table by %IncMSE in descending order
feature_importance <- feature_importance[order(-feature_importance$`%IncMSE`), ]

kable(feature_importance, format = "html", digits = 4, 
      col.names = c("Feature", "%IncMSE", "IncNodePurity"),
      caption = "Feature Importance") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:3, width = "150px")

#Test for spatial auto-correlation 

##------------ Create a non-random test set 
# Prepare the data
data <- combined_data_with_rasters %>%
  filter(has_AIS & has_SAR) %>%
  dplyr::select(total_fishing_hours, total_presence_score, lon_std, lat_std, dist_shore, dist_ports, bathy) %>%
  na.omit()

# Split the data into training and test sets
set.seed(123)  # for reproducibility
train_indices <- sample(1:nrow(data), 0.7 * nrow(data))
train_data <- data[train_indices, ]
test_data <- data[-train_indices, ]

# Train the random forest model
#set.seed(123)  # for reproducibility
#rf_timing <- system.time({
#  rf_model_env <- randomForest(
#    total_fishing_hours ~ total_presence_score + lon_std + lat_std + dist_shore + dist_ports + bathy,
#    data = train_data,
#    ntree = 500,
#    importance = TRUE
#  )
#})

#cat("Random Forest model training time:", rf_timing["elapsed"], "seconds\n")

# Save the new model
#rf_model_env_train=rf_model_env
#save(rf_model_env_train, file = "rf_model_01deg_with_rasters_train.Rdata")
load(here::here("rf_model_01deg_with_rasters_train.Rdata"))

# Make predictions on the test set
test_data$predictions <- predict(rf_model_env_train, test_data)

# Calculate residuals
test_data$residuals <- test_data$total_fishing_hours - test_data$predictions

# Prepare spatial data for autocorrelation analysis
test_data_sf <- st_as_sf(test_data, coords = c("lon_std", "lat_std"), crs = 4326)

# Create a neighbors list
k <- 8  # You can adjust this number
nb <- knearneigh(st_coordinates(test_data_sf), k = k)
nb <- knn2nb(nb)

# Create a spatial weights matrix
lw <- nb2listw(nb, style="W")

# Perform Moran's I test on the test set residuals
moran_test <- moran.test(test_data_sf$residuals, lw)

# Create a data frame with the test results
  moran_results <- data.frame(
    Statistic = c("Moran I statistic", "Expectation", "Variance", "Standard deviate", "P-value"),
    Value = c(
      round(moran_test$estimate[1],2),
      round(moran_test$estimate[2],2),
      round(moran_test$estimate[3],2),
      round(moran_test$statistic,2),
      ifelse(moran_test$p.value < 2.2e-16, "< 2.2e-16", format(moran_test$p.value, scientific = TRUE, digits = 4))
    )
  )
  
 # Create and display the table
  kable(moran_results, format = "html", col.names = c("Statistic", "Value"),
        caption = "Moran I Test under randomisation") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE,
                  position = "center") %>%
    column_spec(1, bold = TRUE) %>%
    column_spec(2, width = "15em")
```

### Spatial cross-validation 
```{r}

# Assuming your training_data has lon_std (longitude) and lat_std (latitude) columns
coords <- SpatialPoints(training_data[, c("lon_std", "lat_std")])

# Combine your training data with spatial coordinates
spatial_data <- SpatialPointsDataFrame(coords, data = training_data)

#Testing to split the coordinates with k-means clustering
# Ensure spatial_data is an sf object
if (!inherits(spatial_data, "sf")) {
  spatial_data_sf <- st_as_sf(spatial_data)
} else {
  spatial_data_sf <- spatial_data
}

# Extract coordinates
coords <- st_coordinates(spatial_data_sf)

# Perform k-means clustering
set.seed(123)  # for reproducibility
kmeans_result <- kmeans(coords, centers = 6)

# Add cluster assignments to the sf object
spatial_data_sf$block <- as.factor(kmeans_result$cluster)

# Create the plot
ggplot() +
  geom_sf(data = spatial_data_sf, aes(color = block), size = 1) +
  scale_color_viridis_d(name = "Block") +
  theme_minimal() +
  labs(title = "Spatial Blocks for 6-Fold Cross-Validation",
       subtitle = "Points colored by block assignment") +
  theme(legend.position = "bottom")

# Register cores for parallel processing (use 1 less than your total cores)
#cl <- makeCluster(detectCores() - 1)
#registerDoParallel(cl)

# Initialize a list to store results
#results <- foreach(i = 1:6, .packages = c("randomForest", "dplyr"), .export = "spatial_data_sf") %dopar% {
  
  # Get the indices of the training and test sets for the i-th block
#  test_indices <- spatial_data_sf$block == i
#  train_data <- spatial_data_sf[!test_indices, ]
#  test_data <- spatial_data_sf[test_indices, ]
  
  # Train Random Forest on the training data
#  rf_model <- randomForest(
#    total_fishing_hours ~ total_presence_score + lon_std + lat_std + dist_shore + dist_ports + bathy,
#    data = train_data,
#    ntree = 500,
#    importance = TRUE
#  )
  
  # Save the model
#  model_filename <- paste0("rf_model_fold_", i, ".rds")
#  saveRDS(rf_model, file = model_filename)
  
  # Predict on the test data
#  predictions <- predict(rf_model, test_data)
  
  # Return a list with results and model filename
#  list(
#    results = data.frame(observed = test_data$total_fishing_hours, predicted = predictions, fold = i),
#    model_filename = model_filename
#  )
#}

# Stop the cluster once done
#stopCluster(cl)

# Extract results and model filenames
#results_df <- do.call(rbind, lapply(results, function(x) x$results))
#model_filenames <- sapply(results, function(x) x$model_filename)

# Save results
#save(results_df, file = "spatial_cv_results.RData")
#save(model_filenames, file = "model_filenames.RData")

load("spatial_cv_results.RData")
load("model_filenames.RData")

# Ensure spatial_data and results_df are aligned
#spatial_data <- spatial_data[order(row.names(spatial_data)), ]
#results_df <- results_df[order(row.names(results_df)), ]

# Get all coordinates
#all_coordinates <- coordinates(spatial_data)

# Create neighbor list for the entire dataset
#k <- 30  # Number of nearest neighbors, adjust as needed
#nb_all <- knn2nb(knearneigh(all_coordinates, k = k))

# Create spatial weights for the entire dataset
#lw_all <- nb2listw(nb_all, style="W", zero.policy = TRUE)

# Function to calculate Moran's I
#calculate_morans_i <- function(residuals, indices, lw_all) {
  # Create a logical vector for subsetting
#  subset_vector <- rep(FALSE, length(lw_all$neighbours))
#  subset_vector[indices] <- TRUE
  
  # Subset the weights list for the current fold
#  lw_subset <- subset.listw(lw_all, subset_vector, zero.policy = TRUE)
  
  # Calculate Moran's I
#  moran_result <- moran.test(residuals, lw_subset, zero.policy = TRUE)
#  return(moran_result)
#}

# Calculate Moran's I for each fold in the spatial cross-validation
#morans_i_results <- list()

#for (i in 1:6) {
  # Get the indices for this fold
#  fold_indices <- which(results_df$fold == i)
  
  # Calculate residuals
#  fold_residuals <- results_df$observed[fold_indices] - results_df$predicted[fold_indices]
  
  # Calculate Moran's I for this fold
#  morans_i_results[[i]] <- calculate_morans_i(fold_residuals, fold_indices, lw_all)
  
  # Print progress
#  print(paste("Completed fold", i))
#}

#save(morans_i_results, file="morans_i_results.Rdata")
load(here::here("morans_i_results.Rdata"))

# Print results
#print("Moran's I for each fold in spatial cross-validation:")
#for (i in 1:5) {
#  print(paste("Fold", i))
#  print(morans_i_results[[i]])
#}

# Extract relevant information from Moran's I results
morans_table <- data.frame(
  Fold = 1:6,
  Moran_I = sapply(morans_i_results, function(x) x$estimate[1]),
  Expectation = sapply(morans_i_results, function(x) x$estimate[2]),
  Variance = sapply(morans_i_results, function(x) x$estimate[3]),
  Std_Deviate = sapply(morans_i_results, function(x) x$statistic),
  P_Value = sapply(morans_i_results, function(x) x$p.value)
)

# Format the table
morans_table_formatted <- morans_table %>%
  mutate(
    Moran_I = round(Moran_I, 4),
    Expectation = format(Expectation, scientific = TRUE, digits = 4),
    Variance = format(Variance, scientific = TRUE, digits = 4),
    Std_Deviate = round(Std_Deviate, 2),
    P_Value = ifelse(P_Value < 2.2e-16, "< 2.2e-16", 
                     format(P_Value, scientific = TRUE, digits = 4))
  )

# Create and print the table using kable and kable_styling
kable(morans_table_formatted,
                            col.names = c("Fold", "Moran's I", "Expectation", "Variance", "Std. Deviate", "p-value"),
                            caption = "Moran's I Results for Each Fold in Spatial Cross-Validation",
                            align = c('c', 'r', 'r', 'r', 'r', 'r'),
                            format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:6, width = "150px")
```